---
title: "5 - exercise"
format: 
  html: 
    embed-resources: true
editor: visual
---

We're going to revisit the ISRG financial statements, focusing on the 5 year annual income statement.

### 1. Load libraries

Insert a code chunk that loads the following libraries: `gt`, `gtExtras`, `svglite`, `webshot2`, `tidyverse`, `readxl`, and `janitor`.

```{r}
#| label: load-libraries
#| warning: false
#| message: false

# Professional Tables

# Data Import & Transformation

```

### 2. Import Excel table

As before, let's import the ISRG financials. I've included them in your Files for this project. The file is `5 yr ISRG quarterly statements.xls`. Importantly, we're going to focus on the 5 year annual income statement, which is tab five (5). Be sure to drop the first four rows as before. Create a tibble called `ISRG_income_raw` where the data will be stored.

```{r}
#| label: import-excel

# ISRG_income_raw <- 

```

### 3. Data cleaning

Insert a code chunk that creates a new tibble `ISRG_income_wide`. Clean the names, create an `item_id`, and rename `x1` to `item_name`. Replace any `NA` numeric values with zeros. Create a `trend` column that bundles the five annual columns so we can drop sparklines in the finished table. You can ignore `group_id` and `item_type` for this exercise.

```{r}
#| label: create-wide-tibble

# ISRG_income_wide <- 

```

### 4. Prepping for export to a fancy table

Grab the code from the lesson and modify it. Create an object `ISRG_for_export` from your wide tibble. Change the title to "Annual Income Statement" and the subtitle to "Intuitive Surgical, Inc. 2018-2022." Change the tab spanner and column labels appropriately. Use sparklines. Highlight the `NET INCOME(LOSS)` row.

```{r}
#| label: create-gt-table

# ISRG_for_export <-

```

### 5. Export to PDF and html

Save your table as `ISRG_Income.pdf` and `ISRG_Income.html`.

```{r}
#| label: export-gt
#| eval: false

```
